# Madison Lakes

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache = TRUE,message=FALSE)
library(tidyverse)
```

## Lake Mendota Freezing and Thawing

Lake Mendota is the largest of the four lakes in Madison, Wisconsin.
The University of Wisconsin sits on part of its southern shore.
The lake is over five miles long from east to west and about four miles wide from north to south at its widest point.
The surface area of the lake is about 4000 hectares (15.5 square miles, about 10,000 acres).

Each winter,
Lake Mendota freezes.
Some winters,
there are multiple periods where the lake freezes,
thaws,
and then freezes again.
Due to its proximity to the University of Wisconsin,
the lake has been heavily studied.
Scientists have noted since the 1850s the dates each winter that the Lake Mendota (and other Madison lakes) freeze and thaw.
The Wisconsin State Climatology Office maintains the records,
in recent decades with assistance from the UW-Madison Department of Atmospheric and Oceanic Sciences.

### Criteria for freezing/thawing

Officially,
Lake Mendota is considered to be *open* if less than half the surface is covered by ice and *closed* if more than have of the surface is ice covered.
The determination of dates when the lakes freezes (closes) and thaws (opens) attempts to follow protocols that originated during the middle of the 1800s.
The lake is determined to be closed if it appears that at least 50% of the lake is covered by ice from several vantage points.
For a change in status (open to closed / closed to open) to be deemed official,
it needs to persist until the next day.
There is admittedly some subjectivity in the determinations of the dates.
On Lake Mendota,
there is an additional criterion (see http://www.aos.wisc.edu/~sco/lakes/msn-lakes_instruc.html).

> Determining the opening and closing dates for Lake Mendota is more of a challenge because the length and shape of the lake would require a sufficiently high vantage point that was not readily available to 19th century observers.  Partly because Lake Mendota has a more irregular shoreline, an important secondary criterion applies for that lake: whether one can row a boat between Picnic Point and Maple Bluff.  This rule arose from the era of E. A. Birge and Chancey Juday (according to Reid Bryson, founder of the UW Meteorology Dept., now known as the Dept. of Atmospheric and Oceanic Sciences), because they frequently were out on the lake in a rowboat, and the ice along that line determined if they could transport a case of beer over to their friends in Maple Bluff.

### Map

```{r map, echo=FALSE, fig.height=2, warning=FALSE}
library("ggmap")

dx = 0.07
dy = 0.04
cx = -89.43
cy = 43.11
mad_coords = c(left = cx - dx,
               right = cx + dx,
               bottom = cy - dy,
               top = cy + dy)
map = get_stamenmap(mad_coords, maptype = "watercolor", zoom = 12)
ggmap(map) +
  annotate("text",label="Lake Mendota",x=cx+0.01,y=cy-0.005) +
  theme_void()
```

### Data

The Lake Mendota data is shared on a Department of Atmospheric and Oceanic Sciences website http://www.aos.wisc.edu/~sco/lakes/Mendota-ice.html.
This data is not directly machine readable into a nice format.
The file `lake-mendota-raw.csv` contains the data
from this webpage
formatted by hand to be machine readable
with slight differences in how the information is stored.
Each row represents a period of time where the lake is frozen,
with some winters appearing multiple times.
(When this happens, the winter value is repeated, unlike the web page, where a double quote indicates it is repeated from the previous row.)
Missing data is recorded as NA (instead of ---).
The variable `winter` is a two-year range
while the dates in the closed and open columns contain a month abbreviation and day without the year.
Months from October through December correspond to the first year of the range while those from January through May are the second.
We will not work with the raw data until later in the semester
when we work on cleaning data.

The file `lake-mendota-clean-2020.csv` has been cleaned and transformed where each row represents a single winter instead of a period of freezing.
Several variables are calculated for each winter as specified in the following table.

#### Variables

The CSV file *lake-mendota-clean-2020.csv* has the following variables.

| **Variable** | Description |
| :----- | :---------- |
| *winter* | two-year range of the stary and end of the winter |
| *year1* | the first year of the winter |
| *intervals* | the number of time intervals the lake is closed |
| *days* | the total number of days the lake is closed |
| *first_freeze* | the date the lake first closes with ice |
| *last_thaw* | the date of the last thaw (day after the last closed date) |
| *decade* | the decade (1850, 1860, ...) of the winter |

## Questions

As we analyze this data,
we will be looking for patterns for how aspects of the freezing and thawing of Lake Mendota has changed over time.
Has the duration of the lake being closed with ice changed; if so how?
How has the typical date that the lake first freezes and last thaws changed over time?
Can trends in the data be explained by modelling a straight line, or is a curve better?
By how do individual observations in a given winter vary from what is typical?
What do we predict might happen in future years?
Is there evidence of a changing climate apparent in this data?

