# Madison Police Incident Reports

```{r setup-7, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache=TRUE,message=FALSE)
library(tidyverse)
library(lubridate)
```

## Public Data

The city of Madison police department maintains a public database with summary information from incidents that are deemed to be of potential interest to the public.
The data represents a selection of police incidents from early in 2005 through the present (late September, 2020, as of this writing).
The data is available for free public download from the website
https://data-cityofmadison.opendata.arcgis.com/datasets/police-incident-reports .
The data set is not complete.

> Incidents listed are selected by the Officer In Charge of each shift that may have significant public interest. Incidents listed are not inclusive of all incidents.

This characteristic of the data means it likely is not representative of all police incidents during this time period.

## Variables

There are 11 variables in the data set, shown in the following table.

```{r data, include = FALSE}
police = read_csv("../stat240/data/Police_Incident_Reports_2020-09-28.csv",
                  col_types = cols(
                    IncidentID = col_integer(),
                    IncidentType = col_character(),
                    CaseNumber = col_character(),
                    IncidentDate = col_character(),
                    Suspect = col_character(),
                    Arrested = col_character(),
                    Address = col_character(),
                    Victim = col_character(),
                    Details = col_character(),
                    ReleasedBy = col_character(),
                    DateModified = col_character())) %>% 
  mutate(IncidentDate = ymd_hms(IncidentDate),
         DateModified = ymd_hms(DateModified))

police %>%
  filter(IncidentDate > ymd("2000-01-01") & IncidentDate < today()) %>% 
  summarize(min = min(IncidentDate),
                     max = max(IncidentDate))

foo = tibble(
  name = names(police),
  description = c("a unique identifier in the data set",
                  "there are nearly 100 types",
                  "case number, internal police identifier",
                  "the date and time of the incident",
                  "a description of the suspect or suspects",
                  "a description of the individual or individuals arrested",
                  "address of the incident",
                  "description of the victim or victims; could be one or more people or a business",
                  "verbal description of the incident",
                  "name of the officer or official who released the report",
                  "date report modified, if applicable"))
```

```{r, echo = FALSE}
library(kableExtra)
foo %>%
  kable() %>% 
  kable_styling(full_width = TRUE, position = "center")
```

## Data Challenges

Each of the variables recorded as free text.
The same information may be expressed very differently in different observations, depending on who was doing the recording.
For example, missing data could be left blank and read in by R as NA.
But text such as "n/a", "N/a", "None", "N/A" or others mean the same thing.
Similarly, there are many alternative ways that the police might enter information about the sex, race, or age of victims, suspects, and people arrested.
Examples include: "15-year old male", "adult male", "M/B 18 years old", "2 males, both 20 years old 1 female, 19 years old", "Two white females, 18-20 years of age", and many others.
Some individuals are identified by name and sex might be inferred.

This lack of a universal means to record the data is generally not an issue for an intelligent human reader trying to read data from a small number of incidents,
but presents a significant challenge when trying to extract summary information from the entire data set.
In addition to the same information being stored in many ways,
a further challenge is that there are some obvious errors in dates,
case numbers, and in other variables.
Values from case numbers and other variables for the same observation or other observations near by in the data set can be useful to correct errors
as the incidents are, more or less, in order of time.

## Strings and Regular Expressions

Extracting useful summary data from the police incident report requires the ability to work with strings and regular expressions.
A *string* is a sequence of characters that could be a single character, or a longer piece of text including letters, digits, punctuation, and other symbols.
A *regular expression* is a pattern of characters, some with special meaning, that is used to match parts or all of strings.

As an example, suppose we wanted to identify all cases where the victim was an individual person and then extract the sex, race, and age of the individual when provided.
We can write code that searches for certain patterns in the text of the string of a cell of the data and classifies the outcome based on the pattern identified.
As the variety of patterns is extensive,
code will need to be quite lengthy to handle the multitude of cases.
We will use this data as a means to learn about string commands in the tidyverse package **stringr** and as an introduction to the powerful tool of regular expressions.

## Questions

Here are some motivating questions.

1. Do the variables with dates have accurate information? Can we identify and correct errors?

2. How much data is missing in each variable?

3. Over what time period do the incidents occur?

4. How does the case number relate to the date of the incident? Are there any errors?

5. What types of incidents are reported? Which are the most common?

6. How often are words such as male, female, white, black, Hispanic, and so on used to describe the suspect?
Can we identify how often the suspect is in each of these categories?

7. How often is the suspect a single individual and when is the suspect more than one person?

8. How often is the victim a business?

9. Can we identify the relative frequency that the person arrested is male or female, or the race?

10. Incidents have dates and times. Are certain types of incidents more or less likely to occur at certain times or certain days of the week?

