# Binomial Distributions

\newcommand{\given}{\,\mid\,}
\newcommand{\E}{\mathsf{E}}
\newcommand{\Var}{\mathsf{Var}}
\renewcommand{\prob}{\mathsf{P}}

```{r setup-10, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
library(tidyverse)
source("scripts/viridis.R")
source("scripts/ggprob.R")
```

One of the most important discrete distribution used in statistics
is the *binomial distribution*.
This is the distribution which counts the number of heads in $n$ independent coin tosses where each individual coin toss
has the probability $p$ of being a head.
The same distribution is useful when not tossing coins,
but for example, when taking random samples and counting the number
of individuals in the sample that fall into a particular category.
Or, if assuming that choices made by a chimpanzee in a prosocial choice experiment are independent and all have the same probability,
the number of prosocial choices made.

## The Binomial Probability Mass Function

The probability mass function for the binomial distribution is
$$
\prob(X=x) = \binom{n}{x} p^x(1-p)^{n-x}, \ \text{for } x=0,\ldots,n
$$
where $n$ is a positive integer and $p$ is a probability between 0 and 1.

This probability mass function uses the expression $\binom{n}{x}$,
called a binomial coefficient
$$
\binom{n}{x} = \frac{n!}{x!(n-x)!}
$$
which counts the number of ways to choose $x$ things from $n$.

**Example**:
Assume a small prosocial choice experiment with where $n=3$ is the number of choices the chimpanzee makes and independence among the choices
and let $X$ be the number of prosocial choices made.

The collection of possible sequence from the previous chapter is

$$
\Omega = \left \{ PPP, PPS, PSP, PSS, SPP, SPS, SSP, SSS \right \}
$$

- $\prob(X=0) = \binom{3}{0} p^0(1-p)^3 = 1 \times (1-p)^3$. The sequence SSS is the only one with no prosocial choices and the probability of the specific sequence is $(1-p) \times (1-p) \times (1-p)$.

- $\prob(X=1) = \binom{3}{1} p(1-p)^2 = 3 \times p(1-p)^2$. There are three sequence, PSS, SPS, and SPP with exactly one prosocial choice. The probability of PSS is $p \times (1-p) \times p$, the probability of SPS is $(1-p) \times p \times (1-p)$ and the probability of SSP is $(1-p) \times (1-p) \times p$. Each of the three sequences has exactly the same probability because multiplication does not depend on the order.

- $\prob(X=2) = \binom{3}{2} p^2(1-p) = 3 \times p^2(1-p)$. Just as the case where $X=1$, there are three sequences and each has the same probability.

- $\prob(X=3) = \binom{3}{3} p^3(1-p)^0 = 1 \times p^3$. The sequence PPP is the only sequence with three prosocial choices.

This example explains the format source of the binomial probability mass expression.
Among all sequences of length $n$ with $x$ P characters and $n-x$ S characters:

- $\binom{n}{x}$ is the number of such sequences 
- $p^x(1-p)^{n-x}$ is the probability of any single such sequence

## Mean and Variance

The mean of the binomial distribution is $\E(X) = np$ and the variance is $\Var(X) = n\prob(1-p)$.
These expressions can be challenging to calculate from the definitions.

$$
\E(X) = \sum_{x=0}^n x \binom{n}{x} p^x(1-p)^{n-x}
$$
$$
\Var(X) = \sum_{x=0}^n (x-np)^2 \binom{n}{x} p^x(1-p)^{n-x}
$$

However, like in many instances, we can calculate these expectations using a method other than the definition.
Here, it is most useful to recognize that the binomial random variable
may be derived as the sum of $n$ independent indicator random variables,
each with probability $p$ of being equal to 1.
$$
X = I_1 + \cdots + I_n
$$
Note that
$$
\E(I_i) = 0 \cdot (1-p)  + 1 \cdot p = p
$$
for each of these indicator variables
and
$$
\Var{I_i} = (0-p)^2 (1-p) + (1-p)^2 p = p^2(1-p)+(1-p)^2p = p(1-p)
$$
Using the equations for expected values and variances of sums of random variables,
we see that

$$
\E(X) = \E(I_1) + \cdots + \E(I_n) = \underbrace{p + \cdots + p}_{\text{$n$ times}} = np
$$

and

$$
\Var(X) =\Var(I_1) + \cdots + \Var(I_n) =  \underbrace{p(1-p) + \cdots + p(1-p)}_{\text{$n$ times}} = np(1-p)
$$

## Binomial Calculations Using R

There are several useful functions for working with the binomial distribution in R.
Each of these functions has the root `binom` and has as a prefix one of `p`, `d`, `r`, or `q`.

Here are examples for the $\text{Binomial}(20,0.4)$ distribution.

## Binomial Random Samples

The function `rbinom()` generates *random* binomial variables.
The first argument the number of values to generate;
the second and third arguments are the parameters $n$ and $p$,
but in R these have the argument names `size` and `prob`
where `n` is used as the number of random variables.

Here is a random sample of 5 binomial random variables.
I repeat twice, once using names for the arguments
and once without doing so.

```{r}
rbinom(n=5,size=20,prob=0.4)
rbinom(5,20,0.4)
```

## Binomial Probabilities in R

The function `dbinom()` measures the *density*, which in R for a discrete
random variable we mean the probability of individual outcomes.
You can specify multiple outcomes or just one.

```{r}
## All 21 probabilities rounded to 6 decimal places
round( dbinom(0:20,20,0.4), 6)
## Just \prob(X=8)
dbinom(8,20,0.4)
```

The function `pbinom()` calculate the sum of the probabilities up to the value.
So, `pbinom(6,20,0.4)` calculates $\prob(X \le 6) = \prob(X=0) + \cdots \prob(X=6)$.

```{r}
## \prob(X <= 6)
pbinom(6,20,0.4)
## Another way
sum( dbinom(0:6,20,0.4) )
```

## Binomial Quantiles

The function `qbinom()` locates one or more quantiles from a binomial distribution.
A value $x$ is the `p` quantile of the distribution of the random variable $X$ if both
$\prob(X \le x) \ge p$ and $\prob(X \ge x) \ge 1-p$ are true.

If $X$ were continuous, we could simply find the value $x$ where $\prob(X < x) = p$.
But for discrete distributions like the binomial,
a single value $x$ might be a whole range of quantiles because probability increases in jumps.

Here is a graph of the *cumulative distribution function* (cdf) of the $\text{Binomial}(20,0.4)$ distribution, $F(x) = \prob(X \le x)$.
The function `pbinom()` calculates this cdf.

```{r, echo = FALSE}
dat = tibble(x = seq(-3,23,1),
             y = pbinom(x,20,0.4))

dat2 = tibble(x = c(0,8),
              xend = c(8,8),
              y = c(0.5,0.5),
              yend = c(0.5,0))

ggplot(dat, aes(x=x, y=y)) +
  geom_step(direction = "hv") +
  geom_point(data = filter(dat,x >=0 & x <=20), color = "red") +
  geom_segment(aes(x=x, xend=xend, y=y, yend=yend), data = dat2, color = "blue",
               linetype = "dotted") +
  ylab("probability") +
  theme_minimal() +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0)
```

The median of the $\text{Binomial}(20,0.4)$ distribution is the 0.5 quantile.
We see this from the previous graph as the inverse of the CDF evaluated at 0.5.
But 8 is also the quantile for any probability between $\prob(X \le 8) \approx 0.5956$ and
$\prob(X \le 7) \approx 0.4159$.

```{r}
qbinom(0.5, 20, 0.4)
```

Note that $\prob(X \le 8) = `r round(pbinom(8,20,0.4),4)` \ge 0.5$
and $\prob(X \ge 8) = `r round(1 - pbinom(7,20,0.4),4)` \ge 1 - 0.5$.

We can also calculate all the deciles of this distribution.

```{r}
qbinom(seq(0.1,0.9,0.1), 20, 0.4)
```

Note that 9 is both the 0.6 and 0.7 quantiles of this distribution.
Each probability $p$ between 0 and 1 corresponds to a single quantile $x$,
one of the possible values (except if it hits the CDF exactly for some value and could be anywhere between two possible values along a horizontal step),
but each possible value corresponds to a range of possible probabilities (along a vertical step).

## Graphing Binomial Distributions

The file *ggprob.R* contains the definition of a function `gbinom()`
which is useful for graphing binomial distributions and
is compatible with the **ggplot2** package.
There are several additional functions including `geom_binom_density()`
which you can use to add a binomial pdf to a plot
and functions for many other probability distributions.
Note: the function `gbinom()` is not part of a package and is not in base R.
You need to source the code in *ggprob.R* just as you do with *viridis.R* to be able to use the function.

**Basic Usage**

The first two arguments are $n$ and $p$.
```{r}
gbinom(90, 0.5)
```

An optional `scale` argument, the default of which is FALSE,
modifies the range of possible $x$ values displayed.

```{r}
gbinom(90, 0.5, scale = TRUE)
```

You have better control over the range of values plotted using arguments are `a` and `b`, endpoints of this interval, and `color` which is used for the line segments.

```{r}
gbinom(90, 0.5, a = 60, b = 65, color = "firebrick")
```

You can add a layer with `geom_binom_density()` or other **ggplot2** commands.
Note that arguments to `gbinom()` are not aesthetics, and so need to be repeated
in `geom_binom_density()`.

```{r}
gbinom(90, 0.5, scale = TRUE, color = "gray") +
  geom_binom_density(90, 0.5, a = 50, scale = TRUE, color = "firebrick") +
  ggtitle(paste0("P(X >= 50) = ", round(1 - pbinom(49,90,0.5),4))) +
  theme_bw()
```


